Automated Formal Synthesis of Digital Controllers for State-Space Physical Plants

### 2017-01-25 cav2017/cav2017.tex
updated benchmarks' table

%+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\subsection{Numeric representations and soundness} 
\label{sec:numeric_rep}
%+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++

The models we consider have two sources of error due to numeric representation. 
The first is the numeric error introduced by the fixed-point numbers employed to model the plant, i.e., to represent the plant dynamics $A_d$, $B_d$ and $x_k$. 
The second is the quantization error introduced by the digital controller, 
which performs operations in the domain of fixed-point numbers. 
In this section we formally outline the errors introduced by the fixed-point representation of numbers. 
% 
Let $\mathcal{F}_{\langle I,F \rangle}(x)$ denote a real number $x$ represented in a fixed point domain, with $I$ bits representing the integer part and $F$ bits representing the decimal part. The smallest number that can be represented in this domain is $c_m=2^{-F}$. Any mathematical operations performed at the precision $\mathcal{F}_{\langle I,F \rangle}(x)$ will introduce errors, that can be upper bounded~\cite{DBLP:conf/arith/BrainTRW15}. Further details are in Appendix~\ref{appendix:numerical_errors}.

@@

In the following, we shall disregard these steady-state errors (caused by FWL effects)
 when the stability is ensured by synthesis, and then verify its safety accounting for the finite-precision errors. 
%+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\section{CEGIS of Safe Controllers for LTI Systems} 
\label{sec:CEGARIS} 
%+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++

@@

\newpage
\appendix
%-------------------------------
\section{Stability of closed-loop models}
\label{sec:appendix-stability}
%-------------------------------

@@

%-------------------------------
\subsection{Encoding of Jury's criterion}
%-------------------------------

@@

Jury's criterion provides necessary and sufficient conditions for the eigenvalues 
to be within the unit circle and, consequently, for the system is stable.  
It is a known stability test from the control system literature~\cite{fadali}.  
Given a 
polynomial $P(z)$ of the form
$$
P(z) = a_{0}z^{N} + a_{1}z^{N-1} + ... a_{N-1}z + a_{N} = 0, \;\; a_{0}\neq 0,
$$
Jury's criterion explores solely the coefficients ($a_{0},a_{1},...,a_{N-1}$) 
of $P(z)$ for checking stability, which we encompass as property $\phi_{stability}$, as detailed below.

This study limits itself to explain the encoding of Jury's criterion for formal synthesis.  
For the stability  test procedure, the following matrix 
$M=[m_{ij}]_{(2N-2)\times N}$ is built  from $P(z)$ coefficients:
$$
M=\left[
\begin{matrix}
  V^{(0)} \\
  V^{(1)} \\
  \vdots \\
  V^{(N-2)}
 \end{matrix}
\right]\mbox{,}
$$
\noindent where $V^{(k)}=[v^{(k)}_{ij}]_{2\times N}$ is such that
$$
v^{(0)}_{ij}=\begin{cases} 
a_{j-1}, & \mbox{if } i=1 \\   v^{(0)}_{(1)(N-j+1)}, & \mbox{if } i=2 
\end{cases}\mbox{, and} 
$$
$$
v^{(k)}_{ij}=\begin{cases} 
0, & \mbox{if } j>n-k\\
v^{(k-1)}_{1j}-v^{(k-1)}_{2j}\cdot\frac{v^{(k-1)}_{11}}{v^{(k-1)}_{21}}, & \mbox{if } j\leq n-k  \mbox{ and } i=1 \\
v^{(k)}_{(1)(N-j+1)}, & \mbox{if } j\leq n-k \mbox{ and } i=2 \\
\end{cases} \mbox{,}
$$
\noindent where $k\in\mathbb{Z}$, such that $0<k<N-2$. $P(z)$ is the 
characteristic polynomial of a stable system if and only if the following 
four propositions hold:
\begin{itemize}
\item $R_{1}$: $P(1)>0$;
\item $R_{2}$: $(-1)^{N}P(-1)>0$;
\item $R_{3}$: $\vert{a_{0}}\vert <a_{N}$;
\item $R_{4}$: $m_{11}>0\iff (m_{31}>0)\wedge (m_{51}>0)\wedge \dots \wedge (m_{(2N-3)(1)>0)}$.
\end{itemize}
In conclusion, the stability property $\phi_{stability}$ is encoded as the following formula:
$$
\phi_{stability}\iff \ (R_{1} \wedge R_{2} \wedge R_{3} \wedge R_{4}).
$$

@@

Discretization of an continuous dynamical system turns the ODE into a difference equation, assuming zero-order hold
for the input $\vec{u}$ and continuous integration for the noise $\vec{w}$, to
\begin{align}
\label{eq:discretization}
\vec{x}_{k+1} &= \vec{A}_d\vec{x}_k+\vec{B}_d\vec{u}_k + \vec{w}_k,\\
y_k &= \vec{C}_d \vec{x}_ k + \vec{D}_d \vec{u}_ k. 
\end{align}
with covariance for $\vec{w}_k \sim \mathcal{N}(0,Q_d)$,
where
\begin{align}
\label{eq:discretize}
\vec{A}_d &= e^{\vec{A} T_s} = \mathcal{L}^{-1} { ( s \vec{I} - \vec{A} )^{-1} }_{t = T_s},\\
\vec{B}_d &= \int_{0}^{T_s} e^{\vec{A} t} dt\ \vec{B} = \vec{A}^{-1} ( \vec{A}_d - \vec{I} ) \vec{B},\\
\vec{C}_d &= \vec{C},\\
\vec{D}_d &= \vec{D},\\
Q_d &= \int_{0}^{T_s} e^{\vec{A} t} Q e^{\vec{A}^T t} dt.
\end{align}
and $T_s$ is the sample time. Then
\begin{align*}
x(kT_s)=x_k \wedge y(kT_s) = y_k, \forall k.
\end{align*}
Let $g_d(k)=\mathcal{G}(t,g(t),T)$ be a function that performs the discretization described above, where $g(t)$
represents the continuous dynamics and $g_d(k)$ the corresponding discrete dynamics. 
Let $G(s)=\mathcal{L}(g(t))$ and $G_d(z)=\mathcal{Z}(g_d(k))$ be the corresponding Laplace and Z-transforms
of $g(t)$ and $g_d(k)$. Given this relation, we have 
$$G_d(z)=G(z)|_{z=e^{sT}} : g_d(k)=\mathcal{G}(t,g(t),T) \wedge T < \frac{1}{.5f_s},$$
where $.5f_s$ is the Nyquist frequency of $g(t)$. This last restriction is introduced to avoid the effects of aliasing
which could cause ``phantom poles'' to appear otherwise. 
The eigenvalues of $\vec{A}_d$ corresponding to the poles of $G_d(z)$, and those of $\vec{A}$ corresponding to the poles of $G(s)$ are similarly related.
$\hat{\lambda}_i=e^{-\lambda_iT} : \hat{\lambda}_i \in \sigma(\vec{A}_d) \wedge \lambda_i \in \sigma(\vec{A})$
where $\sigma(\cdot)$ is the spectrum of a matrix.
The following remark is worth mentioning regarding the above.
\begin{remark}
The witnessed maximum amplitude of the discrete signals $x_k,y_k$ may be smaller to that of $x(t),y(t)$ due to synchronism at fraction-frequency sampling.
This means that reasoning about the state-space must consider the effects of this possibility as well of the case of maximal inputs from the continuous specification.
\end{remark}

\subsection{Errors due to numerical representation} \label{appendix:numerical_errors}

Let $\mathcal{F}_{\langle I,F \rangle}(x)$ denote a real number $x$ represented in a fixed point domain, with $I$ bits representing the integer part and $F$ bits representing the decimal part. The smallest number that can be represented in this domain is $c_m=2^{-F}$.
The following approximation errors will arise in mathematical operations and representation:
\begin{enumerate}
\item {\bf Truncation:}
Let $x$ be a real number, and $\mathcal{F}_{\langle I,F \rangle}(x)$ be the same number represented in a fixed-point domain as above. 
Then $\mathcal{F}_{\langle I,F \rangle}(x) = x-\delta_T$ where the error $ \delta_T=x\ \%_{c_m}\ \tilde x$, and $\%_{c_m}$ is the modulus operation performed on the last bit.

So $\delta_T$ is the truncation error and it will propagate across operations.
\item {\bf Rounding:}
The following errors appear in simple operations. Let $c_1, c_2$ and $c_3$ be real numbers, and $\delta_{T1}$ and $\delta_{T2}$ be the truncation errors caused by representing $c_1$ and $c_2$ in the fixed-point domain as above.
\begin{enumerate}
\item Addition/Subtraction: these operations only propagate errors coming from truncation of the operands, namely 
$\mathcal{F}_{\langle I,F \rangle}(c_1) \pm \mathcal{F}_{\langle I,F \rangle}(c_2) = c_3 + \delta_3$ with $|\delta_3| \leq |\delta_{T1}| + |\delta_{T2}|$.
\item Multiplication: $\mathcal{F}_{\langle I,F \rangle}(c_1) * \mathcal{F}_{\langle I,F \rangle}(c_2) =  c_3 + \delta_3$ with $|\delta_3| \leq |\delta_{T1}*\mathcal{F}_{\langle I,F \rangle}(c_2)| + |\delta_{T2}*\mathcal{F}_{\langle I,F \rangle}(c_1)| + c_m$, where $c_m=2^{-F}$ as above.
\item Division: the operations performed by our controllers in the FWL domain do not include division. However, we do use division in computations at the precision of the plant. $\mathcal{F}_{\langle I,F \rangle}(c_1) / \mathcal{F}_{\langle I,F \rangle}(c_2) =  c_3 + \delta_3$ with $|\delta_3| \leq $, where $c_m=2^{-F}$ as above.
\end{enumerate}

%  Since \eqref{eq:discretization} is a bi-simulation of \eqref{eq:ode}, we may use the semantics in section \ref{sec:model_semantics} to model continuous dynamical systems.

@@

\begin{tabular}[c]{@{}c@{}}Ball\\ Magnetic\\ Levitation\end{tabular} & $\left[\begin{array}{ccc}
1752	& 0			& -34.07\\
0		& 0			& -38.27\\
0 \\ 0 \\ 1.923

>>>>

aazZ%+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\subsection{Numeric representations and soundness} 
\label{sec:numeric_rep}
%+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++

The models we consider have two sources of error due to numeric representation. 
The first is the numeric error introduced by the fixed-point numbers employed to model the plant, i.e., to represent the plant dynamics $A_d$, $B_d$ and $x_k$. 
The second is the quantization error introduced by the digital controller, 
which performs operations in the domain of fixed-point numbers. 
In this section we formally outline the errors introduced by the fixed-point representation of numbers~\cite{DBLP:conf/arith/BrainTRW15}. 
% 
Let $\mathcal{F}_{\langle I,F \rangle}(x)$ denote a real number $x$ represented in a fixed point domain, with $I$ bits representing the integer part and $F$ bits representing the decimal part. The smallest number that can be represented in this domain is $c_m=2^{-F}$.

The following approximation errors will arise:
\begin{enumerate}
\item {\bf Truncation:}
Let $x$ be a real number, and $\mathcal{F}_{\langle I,F \rangle}(x)$ be the same number represented in a fixed-point domain as above. 
Then $\mathcal{F}_{\langle I,F \rangle}(x) = x-\delta_T$ where the error $ \delta_T=x\ \%_{c_m}\ \tilde x$, and $\%_{c_m}$ is the modulus operation performed on the last bit.

So $\delta_T$ is the truncation error and it will propagate across operations.
\item {\bf Rounding:}
The following errors appear in simple operations. Let $c_1, c_2$ and $c_3$ be real numbers, and $\delta_{T1}$ and $\delta_{T2}$ be the truncation errors caused by representing $c_1$ and $c_2$ in the fixed-point domain as above.
\begin{enumerate}
\item Addition/Subtraction: these operations only propagate errors coming from truncation of the operands, namely 
$\mathcal{F}_{\langle I,F \rangle}(c_1) \pm \mathcal{F}_{\langle I,F \rangle}(c_2) = c_3 + \delta_3$ with $|\delta_3| \leq |\delta_{T1}| + |\delta_{T2}|$.
\item Multiplication: $\mathcal{F}_{\langle I,F \rangle}(c_1) * \mathcal{F}_{\langle I,F \rangle}(c_2) =  c_3 + \delta_3$ with $|\delta_3| \leq |\delta_{T1}*\mathcal{F}_{\langle I,F \rangle}(c_2)| + |\delta_{T2}*\mathcal{F}_{\langle I,F \rangle}(c_1)| + c_m$, where $c_m=2^{-F}$ as above.
\item Division: the operations performed by our controllers in the FWL domain do not include division, so these bounds are inconsequential to this paper. 
\end{enumerate}

@@

In the following, we shall disregard these steady-state errors (caused by FWL effects)
 when the stability is ensured by synthesis, and then verify its safety accounting for the finite-precision errors.

%+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\section{CEGIS of Safe Controllers for LTI Systems} 
\label{sec:CEGARIS} 
%+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++

@@

\newpage
\appendix
%-------------------------------
\section{Stability of closed-loop systems}
\label{sec:appendix-stability}
%-------------------------------

@@

@@

Jury's criterion provides necessary and sufficient conditions for the eigenvalues 
to be within the unit circle and, consequently, for the system is stable.  
It is a known stability test from the control system literature~\cite{fadali}.  
Given a 
polynomial $P(z)$ of the form
$$
P(z) = a_{0}z^{N} + a_{1}z^{N-1} + ... a_{N-1}z + a_{N} = 0, \;\; a_{0}\neq 0,
$$
Jury's criterion explores solely the coefficients ($a_{0},a_{1},...,a_{N-1}$) 
of $P(z)$ for checking stability, which we encompass as property $\phi_{stability}$.

This study, however, limits itself to 
explain the encoding of Jury's criterion for the formal synthesis purpose. 
For the stability  test procedure, the following Jury matrix 
$M=[m_{ij}]_{(2N-2)\times N}$ is built  from $P(z)$ coefficients:
$$
M=\left[
\begin{matrix}
  V^{(0)} \\
  V^{(1)} \\
  \vdots \\
  V^{(N-2)}
 \end{matrix}
\right]\mbox{,}
$$
\noindent where $V^{(k)}=[v^{(k)}_{ij}]_{2\times N}$ such that
$$
v^{(0)}_{ij}=\begin{cases} 
a_{j-1}, & \mbox{if } i=1 \\   v^{(0)}_{(1)(N-j+1)}, & \mbox{if } i=2 
\end{cases}\mbox{, and} 
$$
$$
v^{(k)}_{ij}=\begin{cases} 
0, & \mbox{if } j>n-k\\
v^{(k-1)}_{1j}-v^{(k-1)}_{2j}\cdot\frac{v^{(k-1)}_{11}}{v^{(k-1)}_{21}}, & \mbox{if } j\leq n-k  \mbox{ and } i=1 \\
v^{(k)}_{(1)(N-j+1)}, & \mbox{if } j\leq n-k \mbox{ and } i=2 \\
\end{cases} \mbox{,}
$$
\noindent where $k\in\mathbb{Z}$, such that $0<k<N-2$. $P(z)$ is the 
characteristic polynomial of a stable system if and only if the following 
four propositions hold:
\begin{itemize}
\item $R_{1}$: $P(1)>0$;
\item $R_{2}$: $(-1)^{N}P(-1)>0$;
\item $R_{3}$: $\vert{a_{0}}\vert <a_{N}$;
\item $R_{4}$: $m_{11}>0\iff (m_{31}>0)\wedge (m_{51}>0)\wedge \dots \wedge (m_{(2N-3)(1)>0)}$.
\end{itemize}

The stability property $\phi_{stability}$ is then encoded as follows:
$$
\phi_{stability}\iff \ (R_{1} \wedge R_{2} \wedge R_{3} \wedge R_{4}).
$$

@@

Discretization of an continuous dynamical system turns the ODE into a difference equation, assuming zero-order hold
for the input $\vec{u}$ and continuous integration for the noise $\vec{w}$, to
\begin{align}
\label{eq:discretization}
\vec{x}_{k+1} &= \vec{A}_d\vec{x}_k+\vec{B}_d\vec{u}_k + \vec{w}_k,\\
y_k &= \vec{C}_d \vec{x}_ k + \vec{D}_d \vec{u}_ k. 
\end{align}
with covariance for $\vec{w}_k \sim \mathcal{N}(0,Q_d)$,
where
\begin{align}
\label{eq:discretize}
\vec{A}_d &= e^{\vec{A} T_s} = \mathcal{L}^{-1} { ( s \vec{I} - \vec{A} )^{-1} }_{t = T_s},\\
\vec{B}_d &= \int_{0}^{T_s} e^{\vec{A} t} dt\ \vec{B} = \vec{A}^{-1} ( \vec{A}_d - \vec{I} ) \vec{B},\\
\vec{C}_d &= \vec{C},\\
\vec{D}_d &= \vec{D},\\
Q_d &= \int_{0}^{T_s} e^{\vec{A} t} Q e^{\vec{A}^T t} dt.
\end{align}
and $T_s$ is the sample time. Then
\begin{align*}
x(kT_s)=x_k \wedge y(kT_s) = y_k, \forall k.
\end{align*}
Let $g_d(k)=\mathcal{G}(t,g(t),T)$ be a function that performs the discretization described above, where $g(t)$
represents the continuous dynamics and $g_d(k)$ the corresponding discrete dynamics. 
Let $G(s)=\mathcal{L}(g(t))$ and $G_d(z)=\mathcal{Z}(g_d(k))$ be the corresponding Laplace and Z-transforms
of $g(t)$ and $g_d(k)$. Given this relation, we have 
$$G_d(z)=G(z)|_{z=e^{sT}} : g_d(k)=\mathcal{G}(t,g(t),T) \wedge T < \frac{1}{.5f_s},$$
where $.5f_s$ is the Nyquist frequency of $g(t)$. This last restriction is introduced to avoid the effects of aliasing
which could cause ``phantom poles'' to appear otherwise. 
The eigenvalues of $\vec{A}_d$ corresponding to the poles of $G_d(z)$, and those of $\vec{A}$ corresponding to the poles of $G(s)$ are similarly related.
$\hat{\lambda}_i=e^{-\lambda_iT} : \hat{\lambda}_i \in \sigma(\vec{A}_d) \wedge \lambda_i \in \sigma(\vec{A})$
where $\sigma(\cdot)$ is the spectrum of a matrix.
The following remark is worth mentioning regarding the above.
\begin{remark}
The witnessed maximum amplitude of the discrete signals $x_k,y_k$ may be smaller to that of $x(t),y(t)$ due to synchronism at fraction-frequency sampling.
This means that reasoning about the state-space must consider the effects of this possibility as well of the case of maximal inputs from the continuous specification.
\end{remark}
%  Since \eqref{eq:discretization} is a bi-simulation of \eqref{eq:ode}, we may use the semantics in section \ref{sec:model_semantics} to model continuous dynamical systems.

@@

%\begin{tabular}[c]{@{}c@{}}Ball\\ Magnetic\\ Levitation\end{tabular} & $\left[\begin{array}{ccc}
%0		& 1 		& 0		\\
%1752	& 0			& -34.07\\
%0		& 0			& -38.27\\
%\end{array}\right]$ & $\left[\begin{array}{c}
%0 \\ 0 \\ 1.923
%\end{array}\right]$ & $\left[\begin{array}{ccc}
%1 & 0 & 0\\
%\end{array}\right]$ & 0 \\ \hline
